---
title: "Data Manipulation"
author: "Sarahy Martinez"
output: github_document
---

```{r}
library(tidyverse)  # a collection of packages that includes readr,dyplr etc

```

# Load the Litters dataset 

```{r}
litters_df = read_csv("./data/FAS_litters.csv") 

#now update the variables name 

litters_df = janitor :: clean_names(litters_df)

```

## `select`
Choose some column and not others

```{r}

select(litters_df, group, gd0_weight)

```



```{r}
select(litters_df, group, gd0_weight:gd_of_birth)

#lets you select a range 

```


```{r}
# Can also specify what we want to lose 

select(litters_df, -litter_number)

```


Renaming colunms

```{r}
select(litters_df, GROUP = group, LITTer_NUmber = litter_number) 
# can use select to identift columns we want to keep and rename 

```

Instead of selecting and renaming, and just want to rename 

```{r}
rename( litters_df, GROUP = group, LITTer_NUmber = litter_number)
```


Function used specifically to help but wont use entirely 

Select Helpers 
- functions selecting columns knowing we want to keep

```{r}
select(litters_df, starts_with("gd")) #helps select everything we want to start with gd
```


Want to select litter number and not delete everything 

```{r}
select(litters_df, litter_number, everything())
```

Key function we want to use is select, but there are other functions that duplicate 

```{r}
relocate(litters_df, litter_number)
```










