---
title: "Data Manipulation"
author: "Sarahy Martinez"
output: github_document
---

```{r}
library(tidyverse)  # a collection of packages that includes readr,dyplr etc

```

# Load the Litters dataset 

```{r}
litters_df = read_csv("./data/FAS_litters.csv", na = c("NA", ".", ""))

#now update the variables name 

litters_df = janitor :: clean_names(litters_df)

```

## `select`
Choose some column and not others

```{r}

select(litters_df,group, gd0_weight)

```



```{r}
select(litters_df, group, gd0_weight:gd_of_birth)

#lets you select a range 

```


```{r}
# Can also specify what we want to lose 

select(litters_df, -litter_number)

```


#Renaming colunms

```{r}
select(litters_df, GROUP = group, LITTer_NUmber = litter_number) 
# can use select to identift columns we want to keep and rename 

```

Instead of selecting and renaming, and just want to rename 

```{r}
rename( litters_df, GROUP = group, LITTer_NUmber = litter_number)
```


Function used specifically to help but wont use entirely 

Select Helpers 
- functions selecting columns knowing we want to keep

```{r}
select(litters_df, starts_with("gd")) #helps select everything we want to start with gd
```


Want to select litter number and not delete everything 

```{r}
select(litters_df, litter_number, everything())
```

Key function we want to use is select, but there are other functions that duplicate 

```{r}
relocate(litters_df, litter_number)
```


## `filter` 
Intended to remove or keep rows you are interested in retaining for subsequent analysis 
What complicates filter instead of giving variable still need to give it a condition like true or false 

```{r}
filter(litters_df, gd0_weight < 22)
```
We are left with 8 rows instead of original 48

```{r}
filter(litters_df, gd0_weight >= 22)
```

```{r}
filter(litters_df, gd_of_birth == 20) # two equals becuase not trying to make the value go into the variable, this is testing if value 20 and gd birth are equal to eachother, so will give exact value 20
```

There are also ways to negate the things we are interested in

```{r}
filter(litters_df, !(gd_of_birth == 20)) # all pups where not =20
```

Other way 
```{r}
filter(litters_df, gd_of_birth != 20) 

```


```{r}
filter(litters_df, gd0_weight >= 22, gd_of_birth == 20)
```

Instead of numeric, what if we talk about something else 

```{r}
filter(litters_df, group == "Mod8")
```

```{r}
filter(litters_df, group %in% c("Con7","Mod8")) #group equal to either Con7 OR group 8 so use percent sign+ word, says either con7 or mod 8
```

## `mutate`


```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight) 
```

```{r}
mutate(litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```




## `arrange`
to put things in order 
we don't use often bc putting things in order something you do in dataset to make more intuitive and 
doesn't change your data analysis

```{r}

arrange(litters_df, pups_born_alive) #placed pups numerical order, can do alphabet by group
```

```{r}
arrange(litters_df, pups_born_alive, gd0_weight) # first in order by pups born alive and then in order for gestational weight 
```



## ` %>%` pipe 

import -> clean names -> select variables  -> mutate -> filter

```{r}
library(dplyr)

litters_data_raw =read_csv("./data/FAS_litters.csv", na = c("NA", ".", ""))
litters_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected = select(litters_clean_names, -pups_survive)
litters_mutated = mutate(litters_data_selected, wt_gain = gd18_weight - gd0_weight)
litters_mutated = litters_data_selected |>
  mutate(wt_gain = gd18_weight - gd0_weight)

litters_without_missing = drop_na(litters_mutated, gd0_weight)   # only deleting from gd0  
```

Instead USE THE PIPE OPERATOR 

```{r}
litters_df = 
  read_csv("./data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  select(-pups_survive) |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) |> 
  drop_na(wt_gain)

```











