---
title: "Tidy_data"
author: "Sarahy Martinez"
date: "2024-09-25"
output: github_document
---


```{r}
library(tidyverse)
options(tibble.print_min = 5)

```

## `pivot_longer`
This will convert from wide format to longer 

```{r}
#this data is SAS format, so we will use the haven package 
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names()
```

each ID we have 4 diff observation on the BDI score as well as visit, we want to end up with instead of having 4 different columns for BDI score we want for ID to have a column with visit, 1m , 6m, 12m and BDI score observed of each and repeat sex and age. There will be duplicates but thats okay. 


Now go from wide fornmat to long format 
pivot_longer(dataset, columns you need to take from a column by column spread out to wide that we want longer  )
```{r}
pulse_data_tidy = 
  pivot_longer(
    pulse_data, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", #specify range interested in, will end up with two new columns one visit and    
    names_prefix = "bdi_score", #specify range interested in, will end up with two new columns one visit and
    values_to = "bdi")          #values existing in rows, new column names need to exist in new variable
                                # so for this we do names_to, then names_prefix to assign the variable that                                  #will store these values 

pulse_data_tidy
    
```
bdi_score_bl unnecessary we can use mutate and strings but instead we can add name_prefix to will change the 
prefix bdi_score_bl to just _bl



rewrite, combine, and extend (to add a mutate step )
```{r, combined }
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi"
    )%>%  
relocate(id, visit) %>%               #so id and visit are next to each other in order, organized columns
  mutate( visit = recode(visit, "bl"= "00m")) # change visit column so bl value is 00m. 
  
```

## `pivot_wider` 
we use this when 

pivot_longer used  column names that contains info that are variables and we need to go from wide to long

We will make up data!
```{r}
analysis_result =
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time= c("pre", " post", "pre", "post"),
    mean= c(4, 8, 3.5,4)
  )

#
analysis_result %>% 
  pivot_wider(                   #just need to tell where names and values, take values in column and make                                     columns out of that
    names_from = "time",
    values_from = "mean"
    
  )

```


What happens if we have data in multiple tables and we want to stack the rows up?

## Binding Rows 
using LOTR data

Step 1: Import each table, read in different subs by reading in different tables in excel

```{r}

fellowship_ring = 
  readxl:: read_excel("./data/LotR_Words.xlsx", range = "B3:D6") %>% 
  # this alone is an issue, missing movie title, so mutate
mutate(movie = "fellowship_ring") # when run here not at all tidy , but read in remaining data sets

two_towers = 
  readxl:: read_excel("./data/LotR_Words.xlsx", range = "F3:H6") %>% 
mutate(movie = "two_towers") 

return_king = 
  readxl:: read_excel("./data/LotR_Words.xlsx", range = "J3:L6") %>% 
mutate(movie = "return_king") 


```


Bind all rows together
```{r}

lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>%  #combined but not tidy, need to clean names etc.
  janitor::clean_names() %>% 
  relocate(movie) %>% 
  pivot_longer(
    female:male,
    names_to = "gender",
    values_to = "words"  #assigned values that were gender race to words 
  )

```

## Joining Dataset
lets say they share a key 

Import the FAS datasets 

```{r}
pups_df= 
  read_csv("./data/FAS_pups.csv")

litters_df= 
 read_csv("./data/FAS_litters.csv")
```


If we want to know in the pup level data can I tell you something about what happens in the control 7 group?
- I will need to get cntrl 7 group and other things from Litters to pups, merge litter to pups 
litter number is key that exist in both places and from pups pull it 

```{r}
pups_df= 
  read_csv("./data/FAS_pups.csv") %>% 
  janitor::clean_names() %>% 
  mutate(sex = recode(sex, `1` = "male", `2`= "female"))
  

litters_df= 
 read_csv("./data/FAS_litters.csv") %>% 
  janitor::clean_names() %>% 
  relocate(litter_number) %>% 
  separate(group, into = c("dose", "day of tx"),sep = 3) #not mutate, update, just separate exist column                                                             group into two separate things.now sep=3 tells you                                                           where to split 


```


Next up, time to join them! 

```{r}
fas_df = 
  left_join(pups_df, litters_df, by="litter_number") %>%  
  janitor::clean_names() %>% #figure out automatically columns and collections
  arrange(litter_number) %>% 
  relocate(
litter_number, dose, day_of_tx)



```


























